---
title: "Analysing pollutant impact on soil functions with Bayesian Networks"
format: html
editor: visual
---

```{r Load packages, include=FALSE}
pkg <- c("bnlearn", "tidyverse")

for (p in pkg) {
  if (!requireNamespace(p, quietly = TRUE)) {
    install.packages(p)
  }
}

lapply(pkg, library, character.only = T)
```

There are two unique components that when taken together represent **Bayesian Networks (BN)**: **graphical** and **probabilistic**.

# Graphical representation

Relationships between variables in Bayesian nets (BN**)** are represented by arcs. Pairs of variables connected by arcs are **directly** dependent. Variables encoded in the BN are called **nodes**, further, **parent** and child nodes are distinguished to describe the **dependency** between variables. Therefor the pair A ⟶ B encodes dependency of child node B on parent node A. This dependency does not however equal causality. Cyclic dependencies such as A ⟶ B ⟶ A are not allowed, for that reason Bayesian Networks are **acyclic**. Direct and acyclic characteristics are main features of graphical representation of relationships between nodes in Bayesian Networks also called Directed Acyclic Graphs.

Create empty graph with arcs

```{r}
dag <- empty.graph(nodes = c("A", "S", "E", "O", "R", "T"))
arc.set <- matrix(c("A", "E",
                    "S", "E",
                    "E", "O",
                    "E", "R",
                    "O", "T",
                    "R", "T"),
                  byrow = T, 
                  ncol = 2,
                  dimnames = list(NULL, c("from", "to")))
arcs(dag) <- arc.set
```

Recall a product of conditional probabilities

```{r}
modelstring(dag)
```

Recall nodes

```{r}
nodes(dag)
```

Recall arcs

```{r}
arcs(dag)
```

# Probabilistic representation

Variables are assigned with states or levels which are then combined according to dependencies between nodes encoded by arcs in the model. Each combination between variables' states is assigned a probability distribution and all combinations form joint probability distribution.
