---
title: "MSAR Model Implementation for Earthworms (Zinc)"
author: "SEnChemTox"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    number_sections: true
    toc_float:
      collapsed: false
      smooth_scroll: false
    linkcolor: blue
    linestretch: 1.5
    urlcolor: blue
    mainfont: Arial
    fontsize: 11pt
    code_folding: hide
    theme: flatly
  pdf_document:
    toc: true
  word_document:
    toc: true
editor_options:
  markdown:
    wrap: 72
---

```{r message=FALSE, warning=FALSE, include=FALSE}
pkg <- c("tidyverse", "drc", "mc2d", "readxl", "fs", "kableExtra", "knitr", "gt", "stringi", "data.table", "lubridate", "openxlsx", "patchwork")

for (i in pkg) {
  if (!requireNamespace(i, quietly = TRUE)) {
    install.packages(i)
  }
}

lapply(pkg, library, character.only = T)
```

This document provides a comprehensive implementation and evaluation of
the Mean Species Abundance Relationship (MSAR) model for assessing the
ecological risk of Zinc to earthworm populations. The documentation
follows the TRACE framework (TRAnsparent and Comprehensive model
Evaludation; @grimm2014trace) to ensure transparency, reproducibility,
and credibility in ecological risk assessment. The MSAR approach
represents an advancement over traditional Species Sensitivity
Distribution (SSD) methods by incorporating full exposure-response
curves rather than single-point estimates (e.g., EC10 or NOEC), thereby
improving ecological RA and maintaining practical applicability for
lower-tier risk assessment (@hoeks2020mean).

### Problem Formulation {.tabset .tabset-fade .tabset-pills}

#### Decision-Making Context

Chemical pollution of soil ecosystems represents a critical driver of
biodiversity loss, yet traditional risk assessment methods based on
Species Sensitivity Distributions (SSDs) fail to capture changes in
species abundance—a key endpoint in biological conservation
(@hoeks2020mean). The MSAR model addresses this gap by translating
laboratory toxicity data into community-level abundance metrics suitable
for regulatory decision-making under frameworks such as REACH and EU
Regulation 793/93/EC.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Create problem formulation summary table using gt
problem_formulation_df <- data.frame(
  Aspect = c("Application", "Stakeholders", "Regulatory Framework", "Model Type"),
  Description = c(
    "Lower-tier ecological risk assessment for soil contaminants",
    "Regulatory agencies (ECHA), risk assessors, environmental consultants",
    "EU Regulation 793/93/EC, REACH, OECD Guidelines 207/222",
    "Population-level effect model bridging individual toxicity to community metrics"
  )
)

gt(problem_formulation_df) %>%
  tab_header(
    title = "Problem Formulation Summary",
    subtitle = "Decision-making context for MSAR model application"
  ) %>%
  cols_label(
    Aspect = "Aspect",
    Description = "Description"
  ) %>%
  cols_align(align = "left", columns = c(Aspect, Description)) %>%
  tab_style(
    style = cell_text(weight = "bold", transform = "uppercase"),
    locations = cells_column_labels()
  ) %>%
  tab_style(
    style = cell_borders(sides = "bottom", weight = "2px"),
    locations = cells_column_labels()
  )
```

#### Specific Research Questions

The MSAR model implementation addresses four specific scientific
questions that align with regulatory needs for ecologically relevant
risk assessment endpoints: How does zinc concentration affect mean
earthworm species abundance in soil ecosystems? This question addresses
the fundamental relationship between chemical exposure and
community-level responses, moving beyond single-species endpoints to
ecosystem-level metrics (@hoeks2020mean). What is the HC5 (hazardous
concentration for 5% of species) based on abundance loss? The HC5
represents a regulatory benchmark comparable to SSD-derived values but
with improved ecological interpretability through abundance-based
metrics (@schoenke2025assessing). How do soil properties (pH, organic
matter) modify zinc bioavailability and effects? Soil chemistry
significantly influences metal bioavailability, with pH and organic
matter content being primary determinants of zinc speciation and
earthworm exposure (@lock2001influence). What is the uncertainty in MSAR
predictions? Quantifying uncertainty is essential for regulatory
acceptance and enables risk managers to understand the confidence bounds
around protective concentration estimates (@hoeks2020mean).

#### Domain of Applicability

The MSAR model's domain of applicability defines the boundaries within
which model predictions can be considered reliable for regulatory
decision-making. Following @grimm2014trace recommendations for
transparent model documentation, we specify:

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Create domain of applicability table using gt
domain_df <- data.frame(
  Domain_Aspect = c("Chemical Class", "Organism Group", "Exposure Route", "Time Scale", "Spatial Scale", "Acceptable Extrapolation"),
  Specification = c("Heavy metals (Zn, Cd, Cu)", "Earthworms (Lumbricidae)", "Soil (mg/kg dry weight)", "Chronic (28-56 days)", "Laboratory to field", "Similar soil types, temperate climates"),
  Scientific_Justification = c(
    "Metal toxicity follows similar exposure-response patterns (Hendriks et al. 2005)",
    "Standard test organisms with extensive toxicity databases (OECD 2016)",
    "Primary exposure pathway for soil-dwelling organisms (Spurgeon et al. 1994)",
    "Matches standard OECD test durations (OECD 2016)",
    "Enables regulatory application with bioavailability correction (Lock & Janssen 2003)",
    "Based on available validation data (Schoenke et al. 2025)"
  ),
  stringsAsFactors = FALSE
)

gt(domain_df) %>%
  tab_header(
    title = "Domain of Applicability",
    subtitle = "Boundaries for reliable MSAR model predictions"
  ) %>%
  cols_label(
    Domain_Aspect = "Domain Aspect",
    Specification = "Specification",
    Scientific_Justification = "Scientific Justification"
  ) %>%
  cols_align(align = "left", columns = c(Domain_Aspect, Specification, Scientific_Justification)) %>%
  tab_style(
    style = cell_text(weight = "bold", transform = "uppercase"),
    locations = cells_column_labels()
  ) %>%
  tab_style(
    style = cell_borders(sides = "bottom", weight = "2px"),
    locations = cells_column_labels()
  ) %>%
  # CORRECTED: Use tab_options() for font size instead of cell_text()
  tab_options(
    table.font.size = "small"
  ) %>%
  # Optional: Add source note
  tab_source_note(
    source_note = "Based on Hoeks et al. 2020 and Hendriks et al. 2005"
  )
```

### Model Description {.tabset .tabset-fade .tabset-pills}

#### Theoretical Foundation

The MSAR model integrates individual-level toxicity data with population
growth theory to derive community-level abundance metrics. This approach
represents a middle ground between simple SSD methods (which lack
ecological interpretability) and complex population models (which
require extensive data for parameterization) (@hoeks2020mean). The
theoretical foundation rests on three key ecological principles:

1)  Logistic Population Growth: Populations follow logistic growth
    dynamics where the intrinsic rate of increase (r) and carrying
    capacity (K) determine population trajectories
    (@pielou1969introduction).

2)  Toxicant Effects on Life History: Chemical exposure affects survival
    and reproduction, which together determine lifetime fecundity (R₀)
    and ultimately population abundance (@hendriks2005meta).

3)  Carrying Capacity Proportionality: The reduction in carrying
    capacity is proportional to the reduction in intrinsic rate of
    increase, enabling translation of individual-level effects to
    population-level metrics (@hendriks2005meta).

### Key Equations

The MSAR model comprises six interconnected equations that translate
chemical concentration into mean species abundance. Each equation builds
upon the previous one, creating a mechanistic link from individual
toxicity to community-level responses.

```{r}
# Create equations table using gt
equations_df <- data.frame(
  Equation = c("Eq 1", "Eq 2", "Eq 3", "Eq 4", "Eq 5", "Eq 6"),
  Formula = c(
     "$Y(C) = \\frac{1}{1 + (C/EC50)^\\beta}$",
    "$\\frac{R(C)}{R(0)} = (1 - Y_{rep}) \\times (1 - Y_{surv})$",
    "$r = \\frac{\\ln(R_0)}{T_g}$",
    "$\\frac{r(C)}{r(0)} = \\frac{\\ln(R(C))}{\\ln(R(0))}$",
    "$\\frac{K(C)}{K(0)} = \\frac{r(C)}{r(0)}$",
    "$MSA(C) = \\frac{1}{n} \\sum_{i=1}^{n} \\frac{K_i(C)}{K_i(0)}$"
  ),
  Description = c(
    "Exposure-Response (Log-Logistic)",
    "Lifetime Fecundity Reduction",
    "Intrinsic Rate of Increase",
    "Rate Reduction",
    "Carrying Capacity Proportionality",
    "Mean Species Abundance"
  ),
  Reference = c(
    "Hoeks et al. 2020",
    "Hoeks et al. 2020",
    "Hendriks et al. 2005",
    "Hoeks et al. 2020",
    "Hendriks et al. 2005",
    "Hoeks et al. 2020"
  )
)

gt(equations_df) %>%
  tab_header(
    title = "MSAR Model Equations",
    subtitle = "Key equations linking chemical exposure to mean species abundance"
  ) %>%
  cols_label(
    Equation = "Equation",
    Formula = "Formula",
    Description = "Description",
    Reference = "Reference"
  ) %>%
   # Enable Markdown rendering for LaTeX formulas
  fmt_markdown(columns = c(Formula)) %>%
  # Align columns (UPDATED: use c() instead of vars())
  cols_align(align = "center", columns = c(Equation, Formula)) %>%
  cols_align(align = "left", columns = c(Description, Reference)) %>%
  # Style header (UPDATED: use c() instead of vars())
  cols_align(align = "center", columns = c(Equation, Formula)) %>%
  cols_align(align = "left", columns = c(Description, Reference)) %>%
  tab_style(
    style = cell_text(weight = "bold", transform = "uppercase"),
    locations = cells_column_labels()
  ) %>%
  tab_style(
    style = cell_borders(sides = "bottom", weight = "2px"),
    locations = cells_column_labels()
  )
```

Equation 1: Exposure-Response Function The exposure-response
relationship follows a 2-parameter log-logistic model, which is commonly
used to fit sigmoidal exposure-response curves in ecotoxicology
(@gadagkar2015computational). The Hill slope (β) represents the
steepness of the curve at its inflection point, with metals typically
exhibiting flatter slopes (β ≈ -0.36) compared to organic compounds
(@hendriks2005meta).

Equation 2: Lifetime Fecundity Reduction This equation combines survival
and reproduction effects into a single metric of lifetime fecundity
reduction. The multiplicative structure assumes independent effects on
survival and reproduction, which is supported by empirical data for
earthworms exposed to metals (@hendriks2005meta).

Equation 3-5: Population Growth Translation These equations translate
individual-level effects into population-level metrics using classical
population growth theory. The key assumption—that generation time (T_g)
is not affected by contaminants—is supported by laboratory data showing
T_g remains constant up to approximately 0.1-1 times the acute LC50
(@hendriks2005meta).

Equation 6: Mean Species Abundance Aggregation The final equation
aggregates species-specific abundance ratios into a community-level
metric. This approach accounts for both inter- and intraspecies
variation in sensitivity, providing a more ecologically relevant
endpoint than SSD-based methods (@hoeks2020mean).

#### Model Parameters

Model parameters are derived from multiple sources, including direct
extraction from toxicity databases, allometric relationships, and
meta-analyses. Parameters are classified as either directly measured
(from toxicity tests) or estimated (from allometric relationships).

Uncertainty in model parameters is characterized using probability
distributions appropriate for each parameter type. EC50/LC50 values
follow log-normal distributions (concentrations must be positive), while
Hill slopes follow normal distributions centered on fitted values
(@hoeks2020mean). The lifetime fecundity (R₀) follows a log-normal
distribution with mean 4 and standard deviation 0.5, reflecting the
range of 2-11.5 observed across earthworm species (@hendriks2007power).

### Data Evaluation {.tabset .tabset-fade .tabset-pills}

#### Data Sources and Quality Criteria

The MSAR model is parameterized using toxicity data extracted from the
"database heavy metals.xlsx" file (Zn sheet), which contains curated
ecotoxicity data from peer-reviewed literature. This initial dataset
will be complemented with data extracted from the ECOTOX database. Data
quality is assessed following standardized criteria adapted from OECD
guidelines and regulatory guidance documents.

```{r}
# Create data quality criteria table using gt
quality_df <- data.frame(
  Criterion_Category = c("Study Design", "Species Information", "Soil Characterization", "Chemical Information", "Statistical Analysis"),
  Requirements = c(
    "OECD 207, 222, ISO 11268; ≥4 replicates; Control survival >80%",
    "Species clearly identified; Age/size class specified",
    "pH, organic matter, soil type reported",
    "Actual measured concentrations; Purity specified",
    "EC50/LC50 with confidence intervals; Method specified"
  ),
  Implementation = c(
    "Filter by Test_Guideline, Quality_Score columns",
    "Filter by Species_name, Organism_Age columns",
    "Filter out NA in pH, Organic_Matter_percent",
    "Use Concentration_mg_kg (measured values)",
    "Use Lower_CI, Upper_CI, beta_value columns"
  )
)

gt(quality_df) %>%
  tab_header(
    title = "Data Quality Criteria",
    subtitle = "Quality filters applied to toxicity database (adapted from QualCrit sheet)"
  ) %>%
  cols_label(
    Criterion_Category = "Criterion Category",
    Requirements = "Requirements",
    Implementation = "Implementation"
  ) %>%
  cols_align(align = "left", columns = c(Criterion_Category, Requirements, Implementation)) %>%
  tab_style(
    style = cell_text(weight = "bold", transform = "uppercase"),
    locations = cells_column_labels()
  ) %>%
  tab_style(
    style = cell_borders(sides = "bottom", weight = "2px"),
    locations = cells_column_labels()
  ) %>%
  tab_style(
    style = cell_text(size = "small"),
    locations = cells_body()
  )
```

### Data Loading and Cleaning {.tabset .tabset-fade .tabset-pills}

The following R code implements robust data cleaning functions that
handle common issues in ecotoxicity databases, including inconsistent
formatting, missing values, and non-numeric entries. These functions are
vectorized to work efficiently with dplyr pipelines.

```{r}
# Check if file exists
if(!file.exists(paste0(path_home_r(), "/", list.files(path_home_r(), pattern = "database heavy metals.xlsx", recursive = T)))) {
  stop("Please ensure 'database heavy metals.xlsx' is in the working directory.")
}

# Read data
earthworm_data <- paste0(path_home_r(), "/", list.files(path_home_r(), pattern = "database heavy metals.xlsx", recursive = T)) |> read_xlsx(sheet = "Zn")

# Clean column names (make.names converts spaces to dots)
colnames(earthworm_data) <- make.names(colnames(earthworm_data))

# Display original column structure
cat("Original columns:", ncol(earthworm_data), "\n")
cat("Original rows:", nrow(earthworm_data), "\n")
```

The default Hill slope (β = -0.36) for metals is derived from
meta-analysis of concentration-effect relationships (@hendriks2005meta),
which found that metals exhibit flatter exposure-response curves
compared to organic compounds. This default value is applied when
species-specific beta values are unavailable, ensuring conservative
estimates of toxicity.

#### Data Filtering and Processing

Data filtering implements the quality criteria
ensuring adequate ecotoxicity data are used for model
parameterization. The filtering process removes records with missing
critical values, inadequate test durations, or non-standard endpoints.

```{r}
earthworm_processed <- earthworm_data %>%
  rename(Eff_conc = Concentration_mg_kg,
         OM_perc = Organic_Matter_percent) |> 
  dplyr::mutate(
    # Categorize soil type
    Soil_Category = dplyr::case_when(
      grepl("Artificial|OECD", Soil_Type, ignore.case = TRUE) ~ "Artificial",
      grepl("Lufa|LUFA", Soil_Type, ignore.case = TRUE) ~ "Lufa",
      grepl("Natural|Garden|Field|Sandy|Loam|Laterite|Wales|Woburn|Chiltern|BRG|BWZ", Soil_Type, ignore.case = TRUE) ~ "Natural",
      TRUE ~ "Unknown"
    )
  ) %>%
  dplyr::filter(
    !is.na(Eff_conc),
    Eff_conc > 0,
    Test_Duration_Days >= 28,
    Endpoint %in% c("EC50", "LC50"),
    Effect %in% c("Reproduction", "Survival")
  )

cat("=== Data Processing Complete ===\n")
cat("Filtered records:", nrow(earthworm_processed), "\n")
cat("Unique species:", length(unique(earthworm_processed$Species.name)), "\n")
cat("Records removed:", nrow(earthworm_data) - nrow(earthworm_processed), "\n")
```

#### Soil Ecotox Heavy Metals Data Summary

The following table summarizes the quality and coverage of the filtered
toxicity database, providing transparency about the data underlying
model predictions.

```{r}
# Data quality summary table
data_quality_summary <- earthworm_data %>%
  dplyr::summarise(
    Total_Records = n(),
    Unique_Species = n_distinct(Species.name),
    Reproduction_Records = sum(Effect == "Reproduction"),
    Survival_Records = sum(Effect == "Survival"),
    Mean_pH = round(mean(pH, na.rm = TRUE), 2),
    Mean_OM = round(mean(Organic_Matter_percent, na.rm = TRUE), 2),
    Studies_With_CI = sum(!is.na(Lower_CI) & !is.na(Upper_CI))
  )

gt(data_quality_summary) %>%
  tab_header(
    title = "Data Quality Summary",
    subtitle = "Summary statistics for filtered zinc-earthworm toxicity database"
  ) %>%
  cols_label(
    Total_Records = "Total Records",
    Unique_Species = "Unique Species",
    Reproduction_Records = "Reproduction Records",
    Survival_Records = "Survival Records",
    Mean_pH = "Mean pH",
    Mean_OM = "Mean OM (%)",
    Studies_With_CI = "Studies with CI"
  ) %>%
  cols_align(align = "center", columns = everything()) %>%
  tab_style(
    style = cell_text(weight = "bold", transform = "uppercase"),
    locations = cells_column_labels()
  ) %>%
  tab_style(
    style = cell_borders(sides = "bottom", weight = "2px"),
    locations = cells_column_labels()
  )
```

The filtered database contains records for multiple earthworm species,
though coverage varies across ecological groups (epigeic, endogeic,
anecic). Species with both survival and reproduction data provide the
most robust parameterization, while species with only one endpoint
contribute limited information to the MSAR (@hoeks2020mean). The pH
range (approximately 4-8) and organic matter content (approximately
1-20%) cover typical agricultural and natural soils, supporting
extrapolation to field conditions (@lock2001influence).

### Conceptual Model {.tabset .tabset-fade .tabset-pills}

#### Key Assumptions

The MSAR model is based on several simplifying assumptions that balance
ecological realism with practical applicability for regulatory risk
assessment. Each assumption is critically evaluated below with reference
to empirical support and potential limitations.

```{r}
# Create assumptions table using gt
assumptions_df <- data.frame(
  Assumption = c(
    "Logistic population growth",
    "Constant generation time (Tg)",
    "K/r proportionality",
    "No interspecies interactions",
    "Constant bioavailability",
    "R0 = 4 (default)"
  ),
  Justification = c(
    "Standard population ecology theory (Pielou 1969)",
    "Limited data on contaminant effects on Tg (Hendriks 2005)",
    "Empirical support from meta-analysis (Hendriks 2005)",
    "Lower-tier assessment scope (Hoeks 2020)",
    "Data limitations; addressed via correction (Hoeks 2020)",
    "Allometric relationships (Hendriks 2007)"
  ),
  Potential_Impact = c(
    "May not capture complex dynamics",
    "Underestimate if Tg affected by contaminants",
    "Some studies contradict (Underwood 2006)",
    "May underestimate community effects",
    "Addressed via bioavailability correction",
    "Range 2-11.5 considered in uncertainty"
  ),
  Reference = c(
    "Pielou 1969",
    "Hendriks et al. 2005",
    "Hendriks et al. 2005",
    "Hoeks et al. 2020",
    "Hoeks et al. 2020",
    "Hendriks 2007"
  )
)

gt(assumptions_df) %>%
  tab_header(
    title = "Conceptual Model Assumptions",
    subtitle = "Critical evaluation of simplifying assumptions underlying MSAR model design"
  ) %>%
  cols_label(
    Assumption = "Assumption",
    Justification = "Justification",
    Potential_Impact = "Potential Impact",
    Reference = "Reference"
  ) %>%
  cols_align(align = "left", columns = c(Assumption, Justification, Potential_Impact, Reference)) %>%
  tab_style(
    style = cell_text(weight = "bold", transform = "uppercase"),
    locations = cells_column_labels()
  ) %>%
  tab_style(
    style = cell_borders(sides = "bottom", weight = "2px"),
    locations = cells_column_labels()
  ) %>%
  tab_style(
    style = cell_text(size = "small"),
    locations = cells_body()
  )
```

Assumption 1: Logistic Population Growth The assumption of logistic
growth is fundamental to population ecology and provides a tractable
framework for linking individual-level effects to population dynamics
(@pielou1969introduction). While real populations may exhibit more
complex dynamics (e.g., Allee effects, time lags), the logistic model
provides a reasonable approximation for earthworms under stable
environmental conditions.

Assumption 2: Constant Generation Time The assumption that generation
time (T_g) is not affected by contaminants is supported by laboratory
data showing T_g remains fairly constant up to approximately 0.1-1 times
the acute LC50 (@hendriks2005meta). However, this assumption may be
violated for contaminants that specifically affect development rates,
potentially leading to underestimation of population-level effects.

Assumption 3: Carrying Capacity Proportionality The proportionality
between carrying capacity (K) and intrinsic rate of increase (r) is
supported by meta-analysis (@hendriks2005meta) but contradicted by some
studies (@underwood2006variation). This assumption enables translation
of individual-level effects to population metrics without requiring
additional parameters.

Assumption 4: No Interspecies Interactions The exclusion of interspecies
interactions (competition, predation, facilitation) is a deliberate
simplification for lower-tier risk assessment. This assumption may lead
to underestimation of community-level effects, particularly in
species-rich assemblages where indirect effects propagate through food
webs (@delaender2008incorporate).

Assumption 5: Constant Bioavailability The assumption of constant
bioavailability is addressed through the bioavailability correction
model (Section 4.2), which accounts for variation in pH, organic matter,
and soil type. However, temporal changes in bioavailability (e.g.,
aging, weathering) are not captured.

Assumption 6: Default Lifetime Fecundity (R₀ = 4) The default value of
R₀ = 4 is derived from allometric relationships across earthworm species
(@hendriks2007power), with uncertainty characterized by a log-normal
distribution (range 2-11.5). This default enables model application when
species-specific fecundity data are unavailable.

### Modelling {.tabset .tabset-fade .tabset-pills}

#### Bioavailability Correction Model 

Metal bioavailability in soil is controlled by physicochemical
properties that influence metal speciation and organism exposure. The
bioavailability correction model follows empirical relationships
established for earthworms exposed to zinc and other metals
(@lock2001influence; @lock2003influence).

\begin{equation}
f_{bio} = \frac{1}{1 + \alpha_{pH} \times 10^{(pH - pH_{ref})}} \times \frac{1}{1 + \alpha_{OM} \times OM^{\beta_{OM}}} \times f_{soil}
\end{equation}

Where: pH effect: Lower pH increases Zn²⁺ solubility and bioavailability
(α_pH ≈ 1.0) Organic matter effect: Higher OM binds Zn through
complexation, reducing bioavailability (α_OM ≈ 0.3, β_OM ≈ 0.7) Soil
type factor: Artificial/OECD soils serve as reference (f_soil = 1.0);
natural soils typically show reduced bioavailability (f_soil ≈ 0.8-0.9).

The pH effect reflects Zn speciation chemistry, where Zn²⁺ (the
bioavailable form) dominates at low pH while Zn precipitates and
complexes at high pH (@spurgeon1994effects). The organic matter effect
reflects Zn binding to carboxyl and phenolic functional groups in soil
organic matter, reducing the freely dissolved concentration available
for uptake (@lock2001influence).

```{r}
# Bioavailability correction function
calculate_bioavailability_factor <- function(pH, OM_perc, soil_category,
                                                    alpha_pH = 1.0, alpha_OM = 0.3,
                                                    beta_OM = 0.7, pH_ref = 6.0) {
  # Handle NA values
  # if(is.na(pH)) pH <- pH_ref
  # if(is.na(OM_perc)) OM_perc <- 4.0
  
  # pH effect
  f_pH <- 1 / (1 + alpha_pH * 10^(pH - pH_ref))
  
  # OM effect (OM should be percentage, e.g., 4.0 not 0.04)
  f_OM <- 1 / (1 + alpha_OM * (OM_perc)^beta_OM)
  
  # Soil type factor
  f_soil <- dplyr::case_when(
    soil_category == "Artificial" ~ 1.0,
    soil_category == "Lufa" ~ 0.9,
    soil_category == "Natural" ~ 0.8,
    TRUE ~ 0.85
  )
  
  f_bio <- f_pH * f_OM * f_soil
  
  # CONSTRAIN: f_bio should typically be 0.3-1.0 for most soils
  # Minimum 0.2 to avoid extreme bioavailability reduction
  return(pmax(0.2, pmin(1.0, f_bio)))
}

# Apply bioavailability
earthworm_bioava <- earthworm_processed %>%
  dplyr::mutate(
    # Convert OM to percentage if stored as fraction
    OM_perc_corrected = ifelse(OM_perc < 1, OM_perc * 100, OM_perc),
    
    # # Clean pH values (handle ranges like "6.3-6.5")
    # pH_clean = sapply(pH, function(x) {
    #   if(is.na(x)) return(NA)
    #   if(grepl("-", as.character(x))) {
    #     range_vals <- as.numeric(strsplit(as.character(x), "-")[[1]])
    #     return(mean(range_vals, na.rm = TRUE))
    #   }
    #   return(as.numeric(x))
    # })
  ) |> 
  dplyr::mutate(
    f_bio = calculate_bioavailability_factor(
      pH = pH,
      OM_perc = OM_perc_corrected,
      soil_category = Soil_Category
    ),
    Eff_conc_bioava = Eff_conc * f_bio
  ) 

cat("=== Bioavailability Check ===\n")
cat("f_bio range:", round(range(earthworm_bioava$f_bio, na.rm = TRUE), 3), "\n")
cat("Effect concentration original range:", round(range(earthworm_bioava$Eff_conc, na.rm = TRUE), 1), "mg/kg\n")
cat("Effect concentration bioavailable range:", round(range(earthworm_bioava$Eff_conc_bioava, na.rm = TRUE), 1), "mg/kg\n")
cat("Bioavailability reduction factor:", round(mean(earthworm_bioava$Eff_conc / earthworm_bioava$Eff_conc_bioava, na.rm = TRUE), 2), "×\n")
```

The bioavailability function returns values between 0.01 and 1.0,
ensuring bioavailable concentrations remain within realistic bounds.
Default parameter values (α_pH = 1.0, α_OM = 0.3, β_OM = 0.7) are
derived from meta-analysis of earthworm-zinc toxicity studies
(@lock2001influence; @lock2003influence).

#### Core Model Functions

The following R functions implement the six MSAR equations Each function includes explicit type conversion.

```{r}
# Equation 1: Exposure-response function (Hoeks et al. 2020)
calculate_response <- function(C, EC50, beta) {
  C <- as.numeric(C)
  EC50 <- as.numeric(EC50)
  beta <- as.numeric(beta)
  
  if(is.na(EC50) | is.na(beta) | EC50 <= 0) {
    return(NA)
  }
  
  # Handle C = 0 explicitly
  if(C == 0 | C < .Machine$double.eps) {
    return(1.0)
  }
  
  # Use POSITIVE beta for power-form equation
  beta <- abs(beta)
  
  # Hoeks et al. 2020, Equation 1
  response <- 1 / (1 + (C/EC50)^beta)
  
  return(pmax(0, pmin(1, response)))
}

# Equation 2: Lifetime fecundity reduction (Hoeks et al. 2020)
calculate_lifetime_fecundity <- function(C, EC50_rep, beta_rep, LC50_surv, beta_surv) {
  # Survival response (fraction surviving)
  Y_surv <- if(!is.null(LC50_surv) & !is.na(LC50_surv) & LC50_surv > 0) {
    calculate_response(C, LC50_surv, beta_surv)
  } else {
    1.0  # No survival data = assume 100% survival
  }
  
  # Reproduction response (fraction reproducing)
  Y_rep <- if(!is.null(EC50_rep) & !is.na(EC50_rep) & EC50_rep > 0) {
    calculate_response(C, EC50_rep, beta_rep)
  } else {
    1.0  # No reproduction data = assume 100% reproduction
  }
  
  # Hoeks et al. 2020, Equation 2: R(C)/R(0) = Y_rep × Y_surv
  R_ratio <- Y_rep * Y_surv
  
  return(pmax(0, pmin(1, R_ratio)))
}


# Equation 3: Intrinsic rate of increase (Hoeks et al. 2020; Hendriks 2005)
calculate_intrinsic_rate <- function(C, EC50_rep, beta_rep, LC50_surv, beta_surv, R0 = 4) {
  R_ratio <- calculate_lifetime_fecundity(C, EC50_rep, beta_rep, LC50_surv, beta_surv)
  
  # Avoid log(0)
  R_ratio <- pmax(R_ratio, 0.001)
  R0 <- pmax(as.numeric(R0), 2)
  
  # Hoeks et al. 2020, Equation 4
  r_ratio <- log(R0 * R_ratio) / log(R0)
  
  return(pmax(0, pmin(1, r_ratio)))
}

# Equation 4: Carrying capacity (Hendriks 2005)
calculate_carrying_capacity <- function(C, EC50_rep, beta_rep, LC50_surv, beta_surv, R0 = 4) {
  return(calculate_intrinsic_rate(C, EC50_rep, beta_rep, LC50_surv, beta_surv, R0))
}

# Equation 5: MSAR calculation (Hoeks et al. 2020)
calculate_MSAR <- function(C, species_fits, R0 = 4) {
  # CRITICAL: Return 1.0 at C=0
  if(is.numeric(C) && length(C) == 1 && (C == 0 | C < .Machine$double.eps)) {
    return(1.0)
  }
  
  n_species <- length(species_fits)
  K_ratios <- numeric(n_species)
  
  for(i in 1:n_species) {
    fit <- species_fits[[i]]
    
    EC50_rep <- if(!is.null(fit$reproduction)) fit$reproduction$EC50 else NA
    beta_rep <- if(!is.null(fit$reproduction)) fit$reproduction$beta else NA
    LC50_surv <- if(!is.null(fit$survival)) fit$survival$EC50 else NA
    beta_surv <- if(!is.null(fit$survival)) fit$survival$beta else NA
    
    K_ratios[i] <- calculate_carrying_capacity(
      C = as.numeric(C),
      EC50_rep = EC50_rep,
      beta_rep = beta_rep,
      LC50_surv = LC50_surv,
      beta_surv = beta_surv,
      R0 = R0
    )
  }
  
  valid_ratios <- K_ratios[!is.na(K_ratios)]
  if(length(valid_ratios) == 0) {
    return(1.0)
  }
  
  return(mean(valid_ratios))
}
```

#### Species Fits Creation

The following function creates species-specific parameter lists from the
processed ecotoxicity database, organizing reproduction and survival
parameters for each species.

```{r}
create_species_fits_corrected <- function(data) {
  species_list <- unique(data$Species.name)
  species_fits <- list()
  
  for(sp in species_list) {
    sp_data <- data %>% dplyr::filter(Species.name == sp)
    
    # Get reproduction parameters
    rep_data <- sp_data %>% dplyr::filter(Endpoint == "EC50", Effect == "Reproduction")
    rep_params <- NULL
    if(nrow(rep_data) > 0 && !is.na(rep_data$Eff_conc_bioava[1]) && rep_data$Eff_conc_bioava[1] > 0) {
      rep_params <- list(
        EC50 = as.numeric(rep_data$Eff_conc_bioava[1]),
        beta = 0.36,
        lower_ci = as.numeric(rep_data$Lower_CI[1]),
        upper_ci = as.numeric(rep_data$Upper_CI[1])
      )
    }
    
    # Get survival parameters
    surv_data <- sp_data %>% dplyr::filter(Endpoint == "LC50", Effect == "Survival")
    surv_params <- NULL
    if(nrow(surv_data) > 0 && !is.na(surv_data$Eff_conc_bioava[1]) && surv_data$Eff_conc_bioava[1] > 0) {
      surv_params <- list(
        EC50 = as.numeric(surv_data$Eff_conc_bioava[1]),
        beta = 0.36,  
        lower_ci = as.numeric(surv_data$Lower_CI[1]),
        upper_ci = as.numeric(surv_data$Upper_CI[1])
      )
    }
    
    # Include species if they have at least one valid endpoint
    if(!is.null(rep_params) | !is.null(surv_params)) {
      species_fits[[sp]] <- list(
        species = sp,
        reproduction = rep_params,
        survival = surv_params,
        pH = if(nrow(sp_data) > 0 && !is.na(sp_data$pH[1])) sp_data$pH[1] else 6.0,
        organic_matter = if(nrow(sp_data) > 0 && !is.na(sp_data$OM_perc[1])) sp_data$OM_perc[1] else 4.0
      )
    }
  }
  
  return(species_fits)
}

# Create species fits (NO NEED for correct_beta_values anymore)
species_fits <- create_species_fits_corrected(earthworm_bioava)

cat("=== Species Fits Created ===\n")
cat("Total species:", length(species_fits), "\n")

# Count species with each endpoint type
rep_count <- sum(sapply(species_fits, function(f) !is.null(f$reproduction)))
surv_count <- sum(sapply(species_fits, function(f) !is.null(f$survival)))
cat("Species with reproduction:", rep_count, "\n")
cat("Species with survival:", surv_count, "\n")

# Extract all EC50 values for SSD
all_ec50 <- c()
for(i in 1:length(species_fits)) {
  fit <- species_fits[[i]]
  if(!is.null(fit$reproduction) && !is.na(fit$reproduction$EC50)) {
    all_ec50 <- c(all_ec50, fit$reproduction$EC50)
  }
}

cat("\n=== EC50 Values for SSD ===\n")
cat("Number of valid EC50 values:", length(all_ec50), "\n")
cat("EC50 values:", paste(round(all_ec50, 2), collapse = ", "), "\n")
cat("EC50 range:", range(all_ec50), "\n")
```

### Model Output {.tabset .tabset-fade .tabset-pills}

#### MSA

The MSAR curve represents the relationship between zinc concentration
and mean species abundance, aggregating species-specific responses into
a community-level metric. The curve is generated across a concentration
gradient spanning environmentally relevant ranges.

```{r}
# Generate MSAR curve function
generate_MSAR_curve <- function(species_fits, C_range, R0 = 4) {
  C_range <- as.numeric(C_range)
  MSA_values <- sapply(C_range, function(C) calculate_MSAR(C, species_fits, R0))
  
  return(data.frame(
    Concentration = C_range,
    MSA = MSA_values
  ))
}

C_range <- c(0, seq(10, 10000, length.out = 99))
msar_curve <- generate_MSAR_curve(species_fits, C_range, R0 = 4)

# Calculate HC5 and HC10
# calculate_HC <- function(msar_curve, threshold) {
#   hc_index <- which.min(abs(msar_curve$MSA - threshold))
#   return(msar_curve$Concentration[hc_index])
# }
# 
# HC5 <- calculate_HC(msar_curve, 0.95)
# HC10 <- calculate_HC(msar_curve, 0.90)
# 
# cat("HC5 (5% abundance loss):", round(HC5, 2), "mg/kg\n")
# cat("HC10 (10% abundance loss):", round(HC10, 2), "mg/kg\n")
```

As documented by @hoeks2020mean, MSAR-based HC5 values are
systematically higher (factor 3.5-8.9) than SSD-based HC5 values. This
difference reflects the MSAR's use of full exposure-response curves
rather than single-point estimates (EC10 or NOEC), resulting in less
conservative but more ecologically interpretable protective
concentrations (@schoenke2025assessing). The HC5 represents the
concentration at which mean species abundance is reduced by 5% relative
to unexposed conditions. This metric is comparable to SSD-derived HC5
values but incorporates full exposure-response curves rather than
single-point estimates, providing improved ecological interpretability
(@hoeks2020mean).

#### Uncertainty Quantification

The Monte Carlo simulation runs 1000 iterations, sampling parameters
from probability distributions characterized in Section 2.3. For each
iteration, a complete MSAR curve is generated, enabling calculation of
confidence intervals at each concentration point. The 95% confidence
interval (2.5th to 97.5th percentiles) represents the range within which
the true MSAR is expected to fall with 95% probability (@hoeks2020mean).

```{r}
# Monte Carlo uncertainty propagation (Hoeks et al. 2020)
quantify_uncertainty <- function(species_fits, C_range, R0, n_iterations = 1000) {
  n_species <- length(species_fits)
  n_conc <- length(C_range)
  
  MSA_matrix <- matrix(NA, nrow = n_iterations, ncol = n_conc)
  
  for(iter in 1:n_iterations) {
    sampled_fits <- list()
    
    for(i in 1:n_species) {
      fit <- species_fits[[i]]
      sampled_fit <- fit
      
      # Sample EC50 from log-normal distribution (concentrations must be positive)
      if(!is.null(fit$reproduction)) {
        ec50_mean <- fit$reproduction$EC50
        ec50_sd <- if(!is.na(fit$reproduction$lower_ci) & !is.na(fit$reproduction$upper_ci)) {
          (fit$reproduction$upper_ci - fit$reproduction$lower_ci) / 3.92
        } else {
          ec50_mean * 0.3
        }
        ec50_sd <- pmax(ec50_sd, ec50_mean * 0.1)  # Minimum 10% uncertainty
        
        sampled_fit$reproduction$EC50 <- rlnorm(1, 
          meanlog = log(ec50_mean), 
          sdlog = sqrt(log(1 + (ec50_sd/ec50_mean)^2)))
        
        beta_mean <- fit$reproduction$beta
        beta_sd <- abs(beta_mean) * 0.3  # 30% uncertainty on slope
        sampled_fit$reproduction$beta <- rnorm(1, mean = beta_mean, sd = beta_sd)
      }
      
      if(!is.null(fit$survival)) {
        ec50_mean <- fit$survival$EC50
        ec50_sd <- if(!is.na(fit$survival$lower_ci) & !is.na(fit$survival$upper_ci)) {
          (fit$survival$upper_ci - fit$survival$lower_ci) / 3.92
        } else {
          ec50_mean * 0.3
        }
        ec50_sd <- pmax(ec50_sd, ec50_mean * 0.1)
        
        sampled_fit$survival$EC50 <- rlnorm(1, 
          meanlog = log(ec50_mean), 
          sdlog = sqrt(log(1 + (ec50_sd/ec50_mean)^2)))
        
        beta_mean <- fit$survival$beta
        beta_sd <- abs(beta_mean) * 0.3
        sampled_fit$survival$beta <- rnorm(1, mean = beta_mean, sd = beta_sd)
      }
      
      sampled_fits[[i]] <- sampled_fit
    }
    
    # Sample R0 from log-normal (Hendriks 2007)
    R0_sampled <- rlnorm(1, meanlog = log(4), sdlog = 0.5)
    
    MSA_matrix[iter, ] <- sapply(C_range, function(C) {
      calculate_MSAR(C, sampled_fits, R0_sampled)
    })
  }
  
  return(list(
    concentration = C_range,
    MSA_mean = apply(MSA_matrix, 2, mean),
    MSA_lower = apply(MSA_matrix, 2, quantile, probs = 0.025),
    MSA_upper = apply(MSA_matrix, 2, quantile, probs = 0.975),
    all_iterations = MSA_matrix
  ))
}

# Run uncertainty analysis (set seed for reproducibility)
set.seed(123)
uncertainty_results <- quantify_uncertainty(species_fits, C_range, R0 = 4, n_iterations = 1000)
```

#### MSA Curve Visualization

```{r}
# Plot MSAR curve
p1 <- ggplot(msar_curve, aes(x = Concentration, y = MSA)) +
  geom_line(color = "blue", linewidth = 1.2) +
  # geom_hline(yintercept = 0.95, linetype = "dashed", color = "red") +
  # geom_hline(yintercept = 0.90, linetype = "dashed", color = "orange") +
  # geom_vline(xintercept = HC5, linetype = "dashed", color = "red") +
  # geom_vline(xintercept = HC10, linetype = "dashed", color = "orange") +
  scale_x_log10(breaks = c(1, 10, 100, 1000, 10000)) +
  scale_y_continuous(limits = c(0, 1.05)) +
  labs(
    # x = "Zinc Concentration (mg/kg dry soil)",
    y = "MSA",
    title = "MSAR for Earthworms Exposed to Zinc",
    # subtitle = paste("HC5 =", round(HC5, 2), "mg/kg | HC10 =", round(HC10, 2), "mg/kg |", 
    #                  length(species_fits), "species")
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5),
    panel.grid.minor = element_blank()
  )



# Plot MSAR with uncertainty bounds
plot_data <- data.frame(
  Concentration = uncertainty_results$concentration,
  MSA = uncertainty_results$MSA_mean,
  MSA_lower = uncertainty_results$MSA_lower,
  MSA_upper = uncertainty_results$MSA_upper
)

p2 <- ggplot(plot_data, aes(x = Concentration, y = MSA)) +
  geom_line(color = "blue", linewidth = 1.2) +
  geom_ribbon(aes(ymin = MSA_lower, ymax = MSA_upper), alpha = 0.3, fill = "blue") +
  # geom_hline(yintercept = 0.95, linetype = "dashed", color = "red") +
  scale_x_log10(breaks = c(1, 10, 100, 1000, 10000),
                labels = c("1", "10", "100", "1000", "10000")) +
  scale_y_continuous(limits = c(0, 1.05)) +
  labs(
    x = "Zinc Concentration (mg/kg dry soil)",
    y = "MSA",
    title = "MSAR with 95% Confidence Intervals",
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5),
    panel.grid.minor = element_blank()
  )

p1 / p2

```

### Extracting ECOTOX Data {.tabset .tabset-fade .tabset-pills}

Next step is to try to run the MSAR model with data extracted from the
ECOTOX database, which is a large repository of ecotoxicity data. The
following code implements a function to read pipe-delimited files from
the ECOTOX database.

#### LOAD ECOTOX TABLES

First, download ECOTOX database from the USEPA website (https://cfpub.epa.gov/ecotox/), use this green button named "Download ASCII Data". Once it is downloaded on your computer unzip it. Best place to store the files is somewhere in the "Documents". There are two functions which scan for ECOTOX files names. path_home_r() assures that only files placed under the "Documents" are searched for, then list.files() looks for the specific file name in the directory and subdirectories. If the file is not found, a warning is issued and NULL is returned. If the file is found, it is read using fread() with appropriate parameters for pipe-delimited (||) files. Progress messages are printed to the console to indicate loading status and any errors encountered.

```{r}
# Create directory where ECOTOX data should be looked up
ecotox_path <- path_home_r()

# Function to read pipe-delimited files with error handling
read_ecotox_table <- function(filename, show_progress = TRUE) {
  
  filepath <- list.files(ecotox_path, filename, recursive = T, full.names = T)
  
  if (!file.exists(filepath)) {
    warning(paste("File not found:", filename))
    return(NULL)
  }
  
  if (show_progress) {
    cat(sprintf("  Loading %s... ", filename))
  }
  
  tryCatch({
    # Read pipe-delimited file
    df <- fread(
      filepath,
      sep = "|",
      header = TRUE,
      stringsAsFactors = FALSE,
      na.strings = c("", "NA", "N/A"),
      encoding = "Latin-1",
      showProgress = FALSE
    )
    
    if (show_progress) {
      cat(sprintf("✓ (%s rows, %s columns)\n", 
                  format(nrow(df), big.mark = ","), 
                  ncol(df)))
    }
    
    return(df)
    
  }, error = function(e) {
    if (show_progress) {
      cat(sprintf("✗ Error: %s\n", e$message))
    }
    return(NULL)
  })
}

# Load main data tables
tests <- read_ecotox_table("tests.txt")
results <- read_ecotox_table("results.txt")
species <- read_ecotox_table("species.txt")
chemicals <- read_ecotox_table("chemicals.txt")
references <- read_ecotox_table("references.txt")
media_characteristics <- read_ecotox_table("media_characteristics.txt")

# Load lookup tables for decoding
endpoint_codes <- read_ecotox_table("endpoint_codes.txt", show_progress = FALSE)
effect_codes <- read_ecotox_table("effect_codes.txt", show_progress = FALSE)
concentration_unit_codes <- read_ecotox_table("concentration_unit_codes.txt", 
                                               show_progress = FALSE)
duration_unit_codes <- read_ecotox_table("duration_unit_codes.txt", 
                                          show_progress = FALSE)
```

#### IDENTIFY CHEMICAL CAS NUMBERS

```{r}
# Search for Zinc in chemical names
zinc_chemicals <- chemicals %>%
  filter(grepl("zinc|Zinc|ZINC", chemical_name, ignore.case = TRUE))

cat(sprintf("Found %d Zinc-related chemicals:\n", nrow(zinc_chemicals)))
# print(zinc_chemicals %>% 
#         dplyr::select(cas_number, chemical_name) %>% 
#         as.data.frame())

zinc_cas_numbers <- zinc_chemicals$cas_number

cat(sprintf("\nUsing %d CAS numbers for extraction\n\n", length(zinc_cas_numbers)))
```

#### IDENTIFY EARTHWORM SPECIES

```{r}
# Common earthworm species names
earthworm_keywords <- c(
  "Eisenia fetida",
  "Eisenia andrei",
  "Eisenia foetida",  # Alternative spelling
  "Lumbricus terrestris",
  "Lumbricus rubellus",
  "Aporrectodea caliginosa",
  "Allolobophora chlorotica",
  "Dendrobaena",
  "Octolasion"
)

# Create regex pattern
earthworm_pattern <- paste(earthworm_keywords, collapse = "|")

# Filter species table
earthworm_species <- species %>%
  filter(grepl(earthworm_pattern, latin_name, ignore.case = TRUE))

cat(sprintf("Found %d earthworm species in database:\n", nrow(earthworm_species)))

print(earthworm_species %>% 
        dplyr::select(species_number, latin_name, common_name, class, family) %>%
        as.data.frame())

earthworm_species_numbers <- earthworm_species$species_number
```

#### FILTER TESTS FOR ZINC + EARTHWORMS + SOIL

```{r}
# Filter tests
zinc_earthworm_tests <- tests %>% 
  filter(
    test_cas %in% zinc_cas_numbers &
    species_number %in% earthworm_species_numbers &
    organism_habitat %in% c("Soil", "NonSoil")  # Terrestrial studies
  )

cat(sprintf("Found %d tests matching criteria:\n", nrow(zinc_earthworm_tests)))
cat(sprintf("  - Zinc compounds: %d unique CAS\n", 
            n_distinct(zinc_earthworm_tests$test_cas)))
cat(sprintf("  - Earthworm species: %d unique species\n", 
            n_distinct(zinc_earthworm_tests$species_number)))
cat(sprintf("  - References: %d unique studies\n", 
            n_distinct(zinc_earthworm_tests$reference_number)))
```

#### EXTRACT RESULTS (LC50 AND EC50)

```{r}
# Join results with filtered tests
zinc_earthworm_results <- results %>%
  inner_join(zinc_earthworm_tests, by = "test_id") %>%
  # Filter for relevant endpoints
  filter(
    endpoint %in% c("LC50", "EC50", "NOEC", "LOEC", "NOEL", "LOEL")
  )

cat(sprintf("Retrieved %d results\n", nrow(zinc_earthworm_results)))

# Breakdown by endpoint
endpoint_summary <- zinc_earthworm_results %>%
  group_by(endpoint) %>%
  summarise(
    n_records = n(),
    .groups = "drop"
  )

cat("\nEndpoint distribution:\n")
print(endpoint_summary)
```

#### CATEGORIZE ENDPOINTS (MORTALITY VS REPRODUCTION)

```{r}
# For earthworms, use standard test assumptions
zinc_earthworm_results <- zinc_earthworm_results %>%
  mutate(
    # For earthworms, endpoint types are standardized:
    # LC50 = acute mortality (OECD 207, ISO 11268-1)
    # EC50 = reproduction - cocoon production (OECD 222, ISO 11268-2)
    # NOEC/LOEC = typically reproduction
    effect_category = case_when(
      endpoint %in% c("LC50", "LD50", "LT50") ~ "Mortality",
      endpoint %in% c("EC50", "IC50") ~ "Reproduction",
      endpoint %in% c("NOEC", "NOEL") ~ "Reproduction",
      endpoint %in% c("LOEC", "LOEL") ~ "Reproduction",
      endpoint == "MATC" ~ "Reproduction",
      TRUE ~ "Other"
    ),
    
    # Create MSAR-ready endpoint labels
    endpoint_msar = case_when(
      endpoint == "LC50" ~ "LC50_mortality",
      endpoint == "EC50" ~ "EC50_reproduction",
      endpoint == "NOEC" ~ "NOEC_reproduction",
      endpoint == "LOEC" ~ "LOEC_reproduction",
      endpoint == "NOEL" ~ "NOEL_reproduction",
      endpoint == "LOEL" ~ "LOEL_reproduction",
      TRUE ~ paste0(endpoint, "_", effect_category)
    )
  )

# Verify the categorization
# cat("\nVerifying endpoint categorization:\n")

# Show sample records for each endpoint type
# cat("\nSample LC50 records:\n")
# lc50_sample <- zinc_earthworm_results %>%
#   filter(endpoint == "LC50") %>%
#   dplyr::select(endpoint, effect, measurement, endpoint_msar, XXXX_duration_days) %>%
#   head(3)
# print(lc50_sample)

# cat("\nSample EC50 records:\n")
# ec50_sample <- zinc_earthworm_results %>%
#   filter(endpoint == "EC50") %>%
#   dplyr::select(endpoint, effect, measurement, endpoint_msar, XXXX_duration_days) %>%
#   head(3)
# print(ec50_sample)

# Summary by effect category
effect_summary <- zinc_earthworm_results %>%
  group_by(effect_category, endpoint) %>%
  summarise(n_records = n(), .groups = "drop") %>%
  arrange(desc(n_records))

cat("\nEffect category distribution:\n")
print(effect_summary)

# Summary by endpoint_msar
# endpoint_msar_summary <- zinc_earthworm_results |> 
#   group_by(endpoint_msar) |> 
#   summarise(
#     n_records = n(),
#     median_duration_days = median(XXXX_duration_mean, na.rm = TRUE),
#     .groups = "drop"
#   ) |> 
#   arrange(desc(n_records))
# 
# cat("\nMSAR endpoint distribution:\n")
# print(endpoint_msar_summary)
```
#### EXTRACT CONCENTRATION VALUES

```{r}
# Function to extract primary concentration value
extract_concentration <- function(conc_mean, conc_min, conc_max, conc_mean_op) {
  # Use case_when for vectorized if-else logic
  dplyr::case_when(
    !is.na(conc_mean) ~ as.numeric(conc_mean),
    !is.na(conc_min) ~ as.numeric(conc_min),
    !is.na(conc_max) ~ as.numeric(conc_max),
    TRUE ~ NA_real_  # Default case
  )
}

# Extract concentration from conc1 (primary concentration field)
zinc_earthworm_results <- zinc_earthworm_results %>%
  mutate(
    concentration_value = extract_concentration(
      conc1_mean, conc1_min, conc1_max, conc1_mean_op
    ),
    concentration_unit = conc1_unit,
    concentration_type = conc1_type,
    
    # Extract confidence intervals if available
    concentration_lower = as.numeric(conc1_min),
    concentration_upper = as.numeric(conc1_max)
  )

# Check concentration extraction success
conc_extracted <- sum(!is.na(zinc_earthworm_results$concentration_value))
cat(sprintf("Successfully extracted concentration for %d/%d records (%.1f%%)\n",
            conc_extracted,
            nrow(zinc_earthworm_results),
            100 * conc_extracted / nrow(zinc_earthworm_results)))
```

#### STANDARDIZE CONCENTRATION UNITS TO mg/kg

```{r}
# Check current units
unit_distribution <- zinc_earthworm_results %>%
  filter(!is.na(concentration_value)) %>%
  group_by(concentration_unit) %>%
  summarise(n_records = n(), .groups = "drop") %>%
  arrange(desc(n_records))

cat("\nConcentration units found:\n")
print(unit_distribution)

# Conversion function to mg/kg dry soil
convert_to_mg_kg <- function(value, unit) {
  # Handle NA cases
  if_else(is.na(value) | is.na(unit), NA_real_, {
    # Convert to lowercase for matching
    unit_lower <- tolower(trimws(unit))
    
    # Conversion factors with case_when
    value * case_when(
      unit_lower %in% c("mg/kg", "mg kg-1", "mg/kg soil") ~ 1.0,
      unit_lower %in% c("ug/kg", "µg/kg", "ug kg-1") ~ 0.001,
      unit_lower %in% c("g/kg", "g kg-1") ~ 1000.0,
      unit_lower %in% c("ppm") ~ 1.0,
      unit_lower %in% c("mg/l", "mg l-1") ~ 1.0,
      unit_lower %in% c("ug/g", "µg/g") ~ 1.0,
      unit_lower %in% c("mg/g") ~ 1000.0,
      unit_lower %in% c("%", "percent") ~ 10000.0,
      TRUE ~ NA_real_
    )
  })
}

# Apply conversion
zinc_earthworm_results <- zinc_earthworm_results %>%
  mutate(
    concentration_mg_kg = convert_to_mg_kg(concentration_value, concentration_unit),
    concentration_lower_mg_kg = convert_to_mg_kg(concentration_lower, concentration_unit),
    concentration_upper_mg_kg = convert_to_mg_kg(concentration_upper, concentration_unit)
  )

# Check conversion success
converted <- sum(!is.na(zinc_earthworm_results$concentration_mg_kg))
cat(sprintf("\nSuccessfully converted %d/%d records to mg/kg (%.1f%%)\n",
            converted,
            sum(!is.na(zinc_earthworm_results$concentration_value)),
            100 * converted / sum(!is.na(zinc_earthworm_results$concentration_value))))
```

#### EXTRACT TEST DURATION

```{r}
# Extract observation duration from results
zinc_earthworm_results <- zinc_earthworm_results %>%
  mutate(
    duration_value = extract_concentration(
      obs_duration_mean, obs_duration_min, obs_duration_max, obs_duration_mean_op
    ),
    duration_unit = obs_duration_unit
  )

# Convert duration to days
convert_to_days <- function(value, unit) {
  if_else(is.na(value) | is.na(unit), NA_real_, {
    unit_lower <- tolower(trimws(unit))
    
    value * case_when(
      unit_lower %in% c("d", "day", "days") ~ 1.0,
      unit_lower %in% c("h", "hr", "hour", "hours") ~ 1/24,
      unit_lower %in% c("wk", "week", "weeks") ~ 7.0,
      unit_lower %in% c("mo", "month", "months") ~ 30.0,
      unit_lower %in% c("min", "minute", "minutes") ~ 1/(24*60),
      TRUE ~ NA_real_
    )
  })
}

zinc_earthworm_results <- zinc_earthworm_results %>%
  mutate(
    test_duration_days = convert_to_days(duration_value, duration_unit)
  )

# Summary of test durations
duration_summary <- zinc_earthworm_results %>%
  filter(!is.na(test_duration_days)) %>%
  group_by(endpoint_msar) %>%
  summarise(
    n = n(),
    mean_days = mean(test_duration_days, na.rm = TRUE),
    median_days = median(test_duration_days, na.rm = TRUE),
    min_days = min(test_duration_days, na.rm = TRUE),
    max_days = max(test_duration_days, na.rm = TRUE),
    .groups = "drop"
  )

cat("\nTest duration summary by endpoint:\n")
print(duration_summary)
```

#### EXTRACT SOIL PROPERTIES

```{r}
# Join with media characteristics
zinc_earthworm_data <- zinc_earthworm_results %>%
  left_join(media_characteristics, by = "result_id")

# Extract soil pH
zinc_earthworm_data <- zinc_earthworm_data %>%
  mutate(
    soil_ph = extract_concentration(
      media_ph_mean, media_ph_min, media_ph_max, media_ph_mean_op
    )
  )

# Extract organic matter/carbon
zinc_earthworm_data <- zinc_earthworm_data %>%
  mutate(
    soil_om_value = extract_concentration(
      media_org_matter_mean, 
      media_org_matter_min, 
      media_org_matter_max, 
      media_org_matter_mean_op
    ),
    soil_om_unit = media_org_matter_unit,
    soil_om_type = media_org_matter_type,
    
    # Also check for organic carbon
    soil_oc_value = extract_concentration(
      media_org_carbon_mean,
      media_org_carbon_min,
      media_org_carbon_max,
      media_org_carbon_mean_op
    ),
    soil_oc_unit = media_org_carbon_unit
  )
# Standardize to % organic matter
# Common conversion: OC% × 1.724 = OM%
zinc_earthworm_data <- zinc_earthworm_data %>%
  mutate(
    soil_om_percent = case_when(
      !is.na(soil_om_value) & grepl("%|percent", soil_om_unit, ignore.case = TRUE) ~ soil_om_value,
      !is.na(soil_oc_value) & grepl("%|percent", soil_oc_unit, ignore.case = TRUE) ~ soil_oc_value * 1.724,
      !is.na(soil_oc_value) ~ soil_oc_value * 1.724,  # Assume % if no unit
      TRUE ~ NA_real_
    )
  )

# Extract soil texture (sand, silt, clay %)
zinc_earthworm_data <- zinc_earthworm_data %>%
  mutate(
    sand_percent = extract_concentration(
      media_texture_sand_mean,
      media_texture_sand_min,
      media_texture_sand_max,
      media_texture_sand_mean_op
    ),
    silt_percent = extract_concentration(
      media_texture_silt_mean,
      media_texture_silt_min,
      media_texture_silt_max,
      media_texture_silt_mean_op
    ),
    clay_percent = extract_concentration(
      media_texture_clay_mean,
      media_texture_clay_min,
      media_texture_clay_max,
      media_texture_clay_mean_op
    )
  )

# Extract CEC (cation exchange capacity)
zinc_earthworm_data <- zinc_earthworm_data %>%
  mutate(
    soil_cec_value = extract_concentration(
      media_cec_mean,
      media_cec_min,
      media_cec_max,
      media_cec_mean_op
    ),
    soil_cec_unit = media_cec_unit
  )

# Extract soil moisture
zinc_earthworm_data <- zinc_earthworm_data %>%
  mutate(
    soil_moisture_percent = extract_concentration(
      media_moisture_pct_mean,
      media_moisture_pct_min,
      media_moisture_pct_max,
      media_moisture_pct_mean_op
    )
  )

# Soil property completeness
soil_props_summary <- data.frame(
  Property = c("pH", "Organic Matter", "Sand %", "Silt %", "Clay %", "CEC", "Moisture %"),
  Available = c(
    sum(!is.na(zinc_earthworm_data$soil_ph)),
    sum(!is.na(zinc_earthworm_data$soil_om_percent)),
    sum(!is.na(zinc_earthworm_data$sand_percent)),
    sum(!is.na(zinc_earthworm_data$silt_percent)),
    sum(!is.na(zinc_earthworm_data$clay_percent)),
    sum(!is.na(zinc_earthworm_data$soil_cec_value)),
    sum(!is.na(zinc_earthworm_data$soil_moisture_percent))
  ),
  Percent = c(
    100 * sum(!is.na(zinc_earthworm_data$soil_ph)) / nrow(zinc_earthworm_data),
    100 * sum(!is.na(zinc_earthworm_data$soil_om_percent)) / nrow(zinc_earthworm_data),
    100 * sum(!is.na(zinc_earthworm_data$sand_percent)) / nrow(zinc_earthworm_data),
    100 * sum(!is.na(zinc_earthworm_data$silt_percent)) / nrow(zinc_earthworm_data),
    100 * sum(!is.na(zinc_earthworm_data$clay_percent)) / nrow(zinc_earthworm_data),
    100 * sum(!is.na(zinc_earthworm_data$soil_cec_value)) / nrow(zinc_earthworm_data),
    100 * sum(!is.na(zinc_earthworm_data$soil_moisture_percent)) / nrow(zinc_earthworm_data)
  )
)

cat("\nSoil property data availability:\n")
print(soil_props_summary)
```

#### ADD METADATA (SPECIES, CHEMICALS, REFERENCES)

```{r}
# Join species information
zinc_earthworm_data <- zinc_earthworm_data %>%
  left_join(
    species %>% dplyr::select(species_number, latin_name, common_name, 
                      class, family, genus, species),
    by = "species_number"
  ) %>%
  rename(
    species_latin = latin_name,
    species_common = common_name,
    species_class = class,
    species_family = family,
    species_genus = genus,
    species_name = species
  )

# Join chemical information
zinc_earthworm_data <- zinc_earthworm_data %>%
  left_join(
    chemicals %>% dplyr::select(cas_number, chemical_name),
    by = c("test_cas" = "cas_number")
  ) %>%
  rename(chemical_name_full = chemical_name)

# Join reference information
zinc_earthworm_data <- zinc_earthworm_data %>%
  left_join(
    references %>% dplyr::select(reference_number, author, title, 
                         source, publication_year),
    by = "reference_number"
  )
```

#### CREATE MSAR-READY DATASET

```{r}
# Select and rename columns for MSAR model
msar_dataset <- zinc_earthworm_data %>%
  dplyr::select(
    # Identifiers
    result_id,
    test_id,
    reference_number,
    
    # Chemical information
    cas_number = test_cas,
    chemical_name = chemical_name_full,
    
    # Species information
    species_number,
    species_latin,
    species_common,
    species_family,
    
    # Toxicity endpoint
    endpoint,
    endpoint_msar,
    effect,
    effect_category,
    measurement,
    response_site,
    
    # Concentration (standardized to mg/kg)
    concentration_mg_kg,
    concentration_lower_mg_kg,
    concentration_upper_mg_kg,
    concentration_unit_original = concentration_unit,
    concentration_type,
    
    # Test duration
    test_duration_days,
    duration_unit_original = duration_unit,
    
    # Soil properties
    soil_ph,
    soil_om_percent,
    sand_percent,
    silt_percent,
    clay_percent,
    soil_cec_value,
    soil_cec_unit,
    soil_moisture_percent,
    soil_type,
    
    # Test conditions
    test_location,
    test_type,
    exposure_type,
    organism_lifestage,
    organism_age_mean,
    organism_age_unit,
    
    # Reference
    author,
    title,
    source,
    publication_year,
    
    # Quality indicators
    chem_analysis_method,
    significance_code
  ) %>%
  # Filter for usable data (has concentration value)
  filter(!is.na(concentration_mg_kg)) %>%
  # Remove duplicates
  distinct()

cat(sprintf("Created MSAR dataset with %d records\n", nrow(msar_dataset)))
```

#### FILTER FOR KEY ENDPOINTS (LC50 AND EC50 REPRODUCTION)

```{r}
# Check what endpoint_msar values are available
cat("Available endpoint_msar values:\n")
available_endpoints <- msar_dataset %>%
  count(endpoint_msar, sort = TRUE)
print(available_endpoints)

# Primary endpoints for MSAR model
# LC50_mortality: Acute toxicity (typically 14-day test)
# EC50_reproduction: Chronic toxicity (typically 28-56 day test)
msar_primary <- msar_dataset %>%
  filter(
    endpoint_msar %in% c("LC50_mortality", "EC50_reproduction")
  )

cat(sprintf("Primary endpoints dataset: %d records\n", nrow(msar_primary)))
cat(sprintf("  - LC50 (mortality): %d records\n", 
            sum(msar_primary$endpoint_msar == "LC50_mortality")))
cat(sprintf("  - EC50 (reproduction): %d records\n", 
            sum(msar_primary$endpoint_msar == "EC50_reproduction")))


# If we got very few or no records, expand the filter
# if (nrow(msar_primary) < 10) {
#   cat("\n⚠ Warning: Very few records found with strict filter.\n")
#   cat("Expanding to include all LC50 and EC50 endpoints...\n\n")
#   
#   msar_primary <- msar_dataset %>%
#     filter(grepl("^LC50|^EC50", endpoint_msar))
#   
#   cat(sprintf("Expanded dataset: %d records\n", nrow(msar_primary)))
# }

# Summary by endpoint and species
primary_summary <- msar_primary %>%
  group_by(endpoint_msar, species_latin) %>%
  summarise(
    n_records = n(),
    median_conc_mg_kg = median(concentration_mg_kg, na.rm = TRUE),
    mean_conc_mg_kg = mean(concentration_mg_kg, na.rm = TRUE),
    sd_conc_mg_kg = sd(concentration_mg_kg, na.rm = TRUE),
    min_conc_mg_kg = min(concentration_mg_kg, na.rm = TRUE),
    max_conc_mg_kg = max(concentration_mg_kg, na.rm = TRUE),
    median_test_days = median(test_duration_days, na.rm = TRUE),
    has_soil_ph = sum(!is.na(soil_ph)),
    has_soil_om = sum(!is.na(soil_om_percent)),
    .groups = "drop"
  ) %>%
  arrange(endpoint_msar, median_conc_mg_kg)

cat("\nSummary by endpoint and species:\n")
print(primary_summary, n = Inf)

# Check for species with both LC50 and EC50
species_with_both <- primary_summary %>%
  group_by(species_latin) %>%
  summarise(
    has_LC50 = any(endpoint_msar == "LC50_mortality"),
    has_EC50 = any(endpoint_msar == "EC50_reproduction"),
    total_records = sum(n_records),
    .groups = "drop"
  ) %>%
  filter(has_LC50 & has_EC50)

cat("\nSpecies with both LC50 and EC50 data (ideal for MSAR):\n")
if (nrow(species_with_both) > 0) {
  print(species_with_both)
} else {
  cat("  None - you'll need to use species separately or aggregate across species\n")
}
```

#### CALCULATE SUMMARY STATISTICS

```{r}
# Overall statistics
overall_stats <- msar_primary %>%
  group_by(endpoint_msar) %>%
  summarise(
    n_records = n(),
    n_species = n_distinct(species_latin),
    n_studies = n_distinct(reference_number),
    median_conc_mg_kg = median(concentration_mg_kg, na.rm = TRUE),
    mean_conc_mg_kg = mean(concentration_mg_kg, na.rm = TRUE),
    sd_conc_mg_kg = sd(concentration_mg_kg, na.rm = TRUE),
    min_conc_mg_kg = min(concentration_mg_kg, na.rm = TRUE),
    max_conc_mg_kg = max(concentration_mg_kg, na.rm = TRUE),
    geometric_mean = exp(mean(log(concentration_mg_kg), na.rm = TRUE)),
    median_test_duration = median(test_duration_days, na.rm = TRUE),
    .groups = "drop"
  )

cat("\nOverall statistics by endpoint:\n")
print(overall_stats)

# Species-specific statistics
species_stats <- msar_primary %>%
  group_by(species_latin, endpoint_msar) %>%
  summarise(
    n_records = n(),
    n_studies = n_distinct(reference_number),
    median_conc_mg_kg = median(concentration_mg_kg, na.rm = TRUE),
    geometric_mean = exp(mean(log(concentration_mg_kg), na.rm = TRUE)),
    min_conc = min(concentration_mg_kg, na.rm = TRUE),
    max_conc = max(concentration_mg_kg, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(endpoint_msar, median_conc_mg_kg)

cat("\n\nSpecies-specific statistics:\n")
print(species_stats, n = Inf)
```

#### QUALITY SCORING

```{r}
# Quality scoring function
calculate_quality_score <- function(df) {
  df %>%
    mutate(
      quality_score = 0,
      
      # +2 points for standard test guideline
      quality_score = quality_score + case_when(
        grepl("OECD|ISO|EPA|ASTM", test_type, ignore.case = TRUE) ~ 2,
        TRUE ~ 0
      ),
      
      # +1 point for having soil pH
      quality_score = quality_score + ifelse(!is.na(soil_ph), 1, 0),
      
      # +1 point for having organic matter
      quality_score = quality_score + ifelse(!is.na(soil_om_percent), 1, 0),
      
      # +1 point for having soil texture
      quality_score = quality_score + ifelse(
        !is.na(sand_percent) | !is.na(clay_percent), 1, 0
      ),
      
      # +1 point for having confidence intervals
      quality_score = quality_score + ifelse(
        !is.na(concentration_lower_mg_kg) | !is.na(concentration_upper_mg_kg), 1, 0
      ),
      
      # +1 point for peer-reviewed publication
      quality_score = quality_score + ifelse(
        grepl("journal|proceedings", source, ignore.case = TRUE), 1, 0
      ),
      
      # Cap at 7
      quality_score = pmin(quality_score, 7)
    )
}

msar_primary <- calculate_quality_score(msar_primary)

# Quality score distribution
quality_dist <- msar_primary %>%
  group_by(quality_score) %>%
  summarise(n_records = n(), .groups = "drop") %>%
  mutate(percent = 100 * n_records / sum(n_records))

cat("\nQuality score distribution:\n")
print(quality_dist)

cat(sprintf("\nRecords with quality score ≥ 4: %d (%.1f%%)\n",
            sum(msar_primary$quality_score >= 4),
            100 * sum(msar_primary$quality_score >= 4) / nrow(msar_primary)))
```

#### EXPORT RESULTS

```{r}
# Create output directory
output_dir <- "ecotox_extracted_data"
if (!dir.exists(output_dir)) {
  dir.create(output_dir)
}

# Export to Excel with multiple sheets
output_file <- file.path(output_dir, 
                        paste0("zinc_earthworm_msar_data_", 
                               format(Sys.Date(), "%Y%m%d"), 
                               ".xlsx"))
# Create workbook
wb <- createWorkbook()

# Sheet 1: Primary MSAR data
addWorksheet(wb, "MSAR_Primary_Data")
writeData(wb, "MSAR_Primary_Data", msar_primary)

# Sheet 2: All extracted data
addWorksheet(wb, "All_Extracted_Data")
writeData(wb, "All_Extracted_Data", msar_dataset)

# Sheet 3: Overall statistics
addWorksheet(wb, "Overall_Statistics")
writeData(wb, "Overall_Statistics", overall_stats)

# Sheet 4: Species statistics
addWorksheet(wb, "Species_Statistics")
writeData(wb, "Species_Statistics", species_stats)

# Sheet 5: Quality score distribution
addWorksheet(wb, "Quality_Scores")
writeData(wb, "Quality_Scores", quality_dist)

# Sheet 6: Data extraction summary
extraction_summary <- data.frame(
  Metric = c(
    "Total Zinc tests found",
    "Earthworm species",
    "Total results extracted",
    "Results with concentration",
    "LC50 (mortality) records",
    "EC50 (reproduction) records",
    "Records with soil pH",
    "Records with organic matter",
    "Records with texture data",
    "High quality records (score ≥ 4)",
    "Unique references"
  ),
  Value = c(
    nrow(zinc_earthworm_tests),
    n_distinct(msar_dataset$species_latin),
    nrow(zinc_earthworm_data),
    nrow(msar_dataset),
    nrow(msar_primary[msar_primary$endpoint_msar == "LC50_mortality", ]),
    nrow(msar_primary[msar_primary$endpoint_msar == "EC50_reproduction", ]),
    sum(!is.na(msar_primary$soil_ph)),
    sum(!is.na(msar_primary$soil_om_percent)),
    sum(!is.na(msar_primary$sand_percent) | !is.na(msar_primary$clay_percent)),
    sum(msar_primary$quality_score >= 4),
    n_distinct(msar_primary$reference_number)
  )
)

addWorksheet(wb, "Extraction_Summary")
writeData(wb, "Extraction_Summary", extraction_summary)

# Save workbook
saveWorkbook(wb, output_file, overwrite = TRUE)

cat(sprintf("✓ Data exported to: %s\n", output_file))

# Also export primary data as CSV
csv_file <- file.path(output_dir, 
                     paste0("zinc_earthworm_msar_primary_", 
                            format(Sys.Date(), "%Y%m%d"), 
                            ".csv"))
write_csv(msar_primary, csv_file)
cat(sprintf("✓ Primary data exported to: %s\n", csv_file))
```

#### GENERATE VISUALIZATIONS

```{r message=FALSE, warning=FALSE}
# Plot 1: Concentration distribution by endpoint
p1 <- ggplot(msar_primary, aes(x = endpoint_msar, y = concentration_mg_kg)) +
  geom_boxplot(fill = "lightblue", alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.5, color = "darkblue") +
  scale_y_log10() +
  labs(
    title = "Zinc Toxicity to Earthworms",
    subtitle = "Distribution of LC50 and EC50 values",
    x = "Endpoint",
    y = "Concentration (mg/kg soil, log scale)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

# Plot 2: Species sensitivity comparison
if (nrow(species_stats) > 0) {
  p2 <- ggplot(species_stats, 
               aes(x = reorder(species_latin, median_conc_mg_kg), 
                   y = median_conc_mg_kg,
                   fill = endpoint_msar)) +
    geom_col(position = "dodge") +
    scale_y_log10() +
    coord_flip() +
    labs(
      title = "Species Sensitivity to Zinc",
      subtitle = "Median toxicity values by species",
      x = "Species",
      y = "Median Concentration (mg/kg soil, log scale)",
      fill = "Endpoint"
    ) +
    theme_minimal() +
    theme(
      plot.title = element_text(face = "bold", size = 14),
      legend.position = "bottom"
    )
  
  ggsave(file.path(output_dir, "zinc_species_sensitivity.png"), 
         p2, width = 10, height = 6, dpi = 300)
}

# Plot 3: Test duration vs concentration
p3 <- ggplot(msar_primary %>% filter(!is.na(test_duration_days)), 
             aes(x = test_duration_days, y = concentration_mg_kg, 
                 color = endpoint_msar)) +
  geom_point(alpha = 0.6, size = 3) +
  geom_smooth(method = "lm", se = TRUE, alpha = 0.2) +
  scale_y_log10() +
  labs(
    title = "Test Duration vs Toxicity",
    subtitle = "Relationship between exposure duration and observed effects",
    x = "Test Duration (days)",
    y = "Concentration (mg/kg soil, log scale)",
    color = "Endpoint"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    legend.position = "bottom"
  )

# Plot 4: Soil pH influence (if available)
if (sum(!is.na(msar_primary$soil_ph)) > 10) {
  p4 <- ggplot(msar_primary %>% filter(!is.na(soil_ph)), 
               aes(x = soil_ph, y = concentration_mg_kg, 
                   color = endpoint_msar)) +
    geom_point(alpha = 0.6, size = 3) +
    geom_smooth(method = "loess", se = TRUE, alpha = 0.2) +
    scale_y_log10() +
    labs(
      title = "Soil pH Influence on Zinc Toxicity",
      subtitle = "Effect of pH on observed toxicity values",
      x = "Soil pH",
      y = "Concentration (mg/kg soil, log scale)",
      color = "Endpoint"
    ) +
    theme_minimal() +
    theme(
      plot.title = element_text(face = "bold", size = 14),
      legend.position = "bottom"
    )
}

print(p1)
print(p2)
print(p3)
print(p4)
```

### References

Grimm V, et al. 2014. Towards better modelling and decision support:
Documenting model development, testing, and analysis using TRACE.
Ecological Modelling 280:129-139.

Hoeks S, et al. 2020. Mean Species Abundance as a Measure of
Ecotoxicological Risk. Environmental Toxicology and Chemistry
39:2304-2313.

Hendriks AJ, et al. 2005. Meta-analysis of Intrinsic Rates of Increase
and Carrying Capacity of Populations Affected by Toxic and Other
Stressors. Environmental Toxicology and Chemistry 24:2267-2277.

Hendriks AJ. 2007. The Power of Size: A Meta-analysis Reveals
Consistency of Allometric Regressions. Ecological Modelling 205:196-208.

Schoenke VSAM, et al. 2025. Assessing Chronic Effects of Chemical
Pollution on Biodiversity Using Mean Species Abundance Relationships.
Environmental Toxicology and Chemistry 44:1134-1141.

Lock K, Janssen CR. 2001. Influence of Soil Characteristics on Zinc
Toxicity to Earthworms. Environmental Toxicology and Chemistry
20:2204-2209.

Lock K, Janssen CR. 2003. Influence of Soil Characteristics on Zinc
Toxicity to Enchytraeus albidus. Chemosphere 52:1635-1641.

Spurgeon DJ, Hopkin SP. 1994. Effects of Metal-Contaminated Soils on
Earthworms. Ecotoxicology and Environmental Safety 35:86-95.

OECD. 2016. Test No. 222: Earthworm Reproduction Test (Eisenia
fetida/Eisenia andrei). OECD Guidelines for the Testing of Chemicals.

Pielou EC. 1969. An Introduction to Mathematical Ecology. John Wiley &
Sons. Thunnissen NW, et al. 2022. The Impact of Imidacloprid and
Thiacloprid on the Mean Species Abundance in Aquatic Ecosystems. Science
of the Total Environment 822:153626.

Garman ER, et al. 2020. Validation of Bioavailability-Based Toxicity
Models for Metals. Environmental Toxicology and Chemistry 39:101-117.
